---
title: "Simulating trees under the FBD model"
format: html
editor: visual
---

## Introduction

This chapter introduces the Fossilized Birth-Death (FBD) process using forward simulations.

## Simulating the complete BD tree

We start by simulating a tree using the TreeSim package. This package can simulate trees conditioned on age or number of tips. In stratigraphic paleobiology, the timescale of observation is often prescribed (e.g., lifespan of a basin, time interval covered by a section), which is why we focus on simulating conditioned on age.

```{r simple-bd-tree-simulation}
# fix random seed for reproducibility
seed = 42
set.seed(seed)
# model parameters, modify as you see fit
lambda = 2 # speciation rate
mu = 0.3 # extinction rate
t_max = 1 # duration of the simulation
tree = TreeSim::sim.bd.age(age = t_max,
                           numbsim = 1, # only simulate one tree
                           lambda = lambda,
                           mu = mu,
                           mrca = FALSE, # age is time since origin, tree has a root edge
                           complete = TRUE)[[1]] # get the complete tree
# inspect the tree
tree
```

This produces a phylo object, which we can plot:

```{r simple-bd-tree-plot}
plot(tree,
     root.edge = TRUE,
     main = "Birth-death tree")
axis(1)
mtext("Time since origin [Myr]", side = 1, line = 2)
```

Note that this is a complete tree, meaning it contains extinct lineages.

## Simulating fossils along the tree

Now, let's sample some fossil occurrences along the tree according to a specified sampling rate:

```{r simulate-fossil-object}
psi = 5 # fossil sampling rate
fossils = FossilSim::sim.fossils.poisson(rate = psi,
                                         tree = tree,
                                         root.edge = FALSE)
# inspect the fossils
fossils
```

Here you can inspect the generated fossil object, which contains information about species, the corresponding edge in the tree and age in the hmin and hmax column. Be aware that the entries in 'hmin' and 'hmax' are ages, even though the naming suggests that they might correspond to heights (which is confusing).

Now, let's plot these fossils along the tree:

```{r plot-fossils-along-tree-basic}
plot(fossils,
     tree = tree,
     show.tip.label = TRUE,
     rho = 0) # suppress highlighting of extant tips
```

There are plenty of plotting options available in FossilSim, e.g., you can plot the stratigraphic ranges (in the time domain):

```{r plot-fossils-along-tree-ranges}
plot(fossils,
     tree = tree, 
     show.ranges = TRUE,
     rho = 0)
```

So far, we have focused on simulating fossils only. Now let's sample some of the extant tips too by adding them to the fossil object. Here the parameter `rho` specifies what portion of the extant tips are sampled.

```{r simulate-and-plot-fossils-and-extant}
rho = 0.5 # proportion of extant tips that are sampled
fossils2 = FossilSim::sim.extant.samples(fossils = fossils,
                                         tree = tree,
                                         rho = rho)
# plot the output
plot(fossils2,
     tree = tree,
     extant.col = "red",
     show.tip.label = TRUE)
```

If you inspect the newly generated fossil object, you can see the new samples with an of age 0 have been added.

This gives us the option to introduce the concept of the reconstructed or sampled tree:

```{r compare-sampled-tree-with-base-tree}
#| layout-ncol: 2
plot(fossils2,
     tree = tree,
     extant.col = "red",
     reconstructed = FALSE,
     main = "Simulated tree and fossils")
plot(fossils2,
     tree = tree,
     extant.col = "red",
     reconstructed = TRUE,
     main = "Sampled tree")
```

The sampled tree is the tree with all unsampled lineages (extant and extinct) removed, i.e., it represents the tree that we can reconstruct from our observed samples. The sampled tree is the best we can hope to recover from our available data, as we can never recover unsampled lineages as part of a standard phylogenetic inference.

## Tasks and discussion points

1.  Explore the simulation pipeline by modifying the parameters and the random seed. How large is the intrinsic variation in the trees when parameters are fixed? Are there parameters where the true tree and reconstructed/sampled tree differ most/least?

2.  Think of the timescale you are usually on. What parameters work well on this scale?

3.  Sampling rate: What determines the sampling rate on your timescale? Does the sampling rate depend on different drivers on different spatial and temporal scales (e.g., section vs. basin. vs. global scale)? If you put in more effort, could you increase it?

4.  Sampling in the present, fully extinct trees, and `rho`: `rho` states that the extant tips undergo a different sampling regime. If you are working on fully extinct taxa, what is an appropriate choice of `rho` and what counts as the present? What about the case where you observe a fixed time-interval in the past, but taxa potentially survive longer than the observed time interval? How would that work in the inference?

## Advanced topics

These topics are not necessary for the workshop, but if you're interested and have time you can work through them and potentially already start your first inference.

## Exporting trees with sampled ancestors

For comparability with other software packages, fossil objects have to be "baked into" the phylo object as tips at the end of an edge of length 0. We show this for the curious:

```{r bake-fossils-into-tree}
#| layout-ncol: 2
# generate sampled ancestor tree
SAt = FossilSim::SAtree.from.fossils(tree = tree,
                                     fossils = fossils2)
plot(SAt$tree,
     main = "Sampled ancestor tree\n simulated")
# select extant tips based on the fossil object
sampled_extant_tips = SAt$fossils$tip.label[SAt$fossils$hmin == 0]
# generate sampled tree from SAtree
SAt_sampled = FossilSim::sampled.tree.from.combined(tree = SAt$tree,
                                                    sampled_tips = sampled_extant_tips)
plot(SAt_sampled,
     main = "Sampled ancestor tree, sampled")
```

If you compare these trees to the ones above you will see that they correspond the the tree with the fossils plotted along the edges (with the exception of a change in tip labels).

### Simulating morphological data along an FBD tree

Simulation of morphological data along an FBD tree works as before by handing the fossil object to sim.morpho:

```{r simulate-mk-along-fbd-tree}
#| layout-ncol: 2
# simple Mk2 model
n_states = 2
br_rate = 0.5
trait_num = 10
trait_to_display = 2
morpho_mk = MorphSim::sim.morpho(k = n_states,
                        time.tree = tree,
                        br.rates = br_rate,
                        trait.num = trait_num, 
                        variable = FALSE,
                        fossil = fossils2)
MorphSim::plotMorphoGrid(morpho_mk)
plot(morpho_mk,
     trait = trait_to_display)
title(main = paste0("Trait no. ", trait_to_display))
```

Note that `plotMorphoGrid` shows the traits at the (potentially unsampled) extant and extinct tips.

### Inference

With the morphological data simulated, you can then export the simulated data:

```{r export-morpho-and-age-data}
# export character matrix
MorphSim::write.morpho(data = morpho_mk,
                       file = "char_mat_temp.nex",
                       type = "matrix",
                       all = TRUE,
                       reconstructed = TRUE)
# export taxon ages
MorphSim::write.morpho(data = morpho_mk,
                       file = "ages_temp.tsv",
                       type = "ages",
                       all = TRUE,
                       reconstructed = TRUE)
```

With this data at hand, you can now try to recover your simulation setup parameters using the Bayesian tree inference tool of your choice. Example scripts for this are available under "Inference helpers" (chapter 13).

*Below I am doing my own thing*

```{r}
setwd("C:/Users/deardenr/stratworkshop/Lorentz-Workshop/book/day3")

```

Simulating the complete BD tree, have raised extinction rate
```{r}
# fix random seed for reproducibility
seed = 42
set.seed(seed)
# model parameters, modify as you see fit
lambda = 2 # speciation rate
mu = 0.4 # extinction rate
t_max = 1 # duration of the simulation
tree = TreeSim::sim.bd.age(age = t_max,
                           numbsim = 1, # only simulate one tree
                           lambda = lambda,
                           mu = mu,
                           mrca = FALSE, # age is time since origin, tree has a root edge
                           complete = TRUE)[[1]] # get the complete tree
# inspect the tree
tree
```
###Plotting the tree
```{r}
png(filename="FBDsimtree_RPD.png", width=800, height=800)
plot(tree,
     root.edge = TRUE,
     main = "Birth-death tree")
axis(1)
mtext("Time since origin [Myr]", side = 1, line = 2)
dev.off()
```
Simulating fossils: raised sampling rate
```{r}
psi = 10 # fossil sampling rate
fossils = FossilSim::sim.fossils.poisson(rate = psi,
                                         tree = tree,
                                         root.edge = FALSE)
# inspect the fossils
fossils
```
Plot fossils
```{r}
png(filename="FBDsimtreefoss_RPD.png", width=800, height=800)
plot(fossils,
     tree = tree,
     show.tip.label = TRUE,
     rho = 0) # suppress highlighting of extant tips
dev.off()

png(filename="FBDsimtreefossrang_RPD.png", width=800, height=800)
plot(fossils,
     tree = tree, 
     show.ranges = TRUE,
     rho = 0)
dev.off()
```
Sample some extants by adding to fossil object
```{r}
rho = 0.45 # proportion of extant tips that are sampled
fossils2 = FossilSim::sim.extant.samples(fossils = fossils,
                                         tree = tree,
                                         rho = rho)
# plot the output
png(filename="FBDsimtreefossrangext_RPD.png", width=800, height=800)
plot(fossils2,
     tree = tree,
     extant.col = "red",
     show.tip.label = TRUE)
dev.off()

```
Introduce reconstructed tree
```{r}
png(filename="FBDsimtreefossboth1_RPD.png", width=1600, height=800)
plot(fossils2,
     tree = tree,
     extant.col = "red",
     reconstructed = FALSE,
     main = "Simulated tree and fossils")
dev.off()

png(filename="FBDsimtreefossboth2_RPD.png", width=1600, height=800)
plot(fossils2,
     tree = tree,
     extant.col = "red",
     reconstructed = TRUE,
     main = "Sampled tree")
dev.off()
```

Aim to have high speciation and low extinction with a high fossil sampling rate. We are foraminiferanologists.
```{r}
# fix random seed for reproducibility
seed = 8
set.seed(seed)
# model parameters, modify as you see fit
lambda = 5 # speciation rate
mu = 0.1 # extinction rate
t_max = 1 # duration of the simulation
ftree = TreeSim::sim.bd.age(age = t_max,
                           numbsim = 1, # only simulate one tree
                           lambda = lambda,
                           mu = mu,
                           mrca = FALSE, # age is time since origin, tree has a root edge
                           complete = TRUE)[[1]] # get the complete tree
# inspect the tree
png(filename="FBD_ftree_RPD.png", width=800, height=800)
plot(ftree,
     root.edge = TRUE,
     main = "Birth-death tree")
axis(1)
mtext("Time since origin [Myr]", side = 1, line = 2)
dev.off()

psi = 20 # fossil sampling rate
ffossils = FossilSim::sim.fossils.poisson(rate = psi,
                                         tree = ftree,
                                         root.edge = FALSE)
# inspect the fossils
ffossils
png(filename="FBD_ftree_fossls_RPD.png", width=800, height=800)
plot(ffossils,
     tree = ftree,
     show.tip.label = TRUE,
     rho = 0) # suppress highlighting of extant tips
dev.off()

png(filename="FBD_ftree_rang.png", width=800, height=800)
plot(ffossils,
     tree = ftree, 
     show.ranges = TRUE,
     rho = 0)
dev.off()

rho = 0.5 # proportion of extant tips that are sampled
ffossils2 = FossilSim::sim.extant.samples(fossils = ffossils,
                                         tree = ftree,
                                         rho = rho)

# plot the output
png(filename="FBD_ftree_extsamp_RPD.png", width=800, height=800)
plot(ffossils2,
     tree = ftree,
     extant.col = "red",
     show.tip.label = TRUE)
dev.off()

```
