---
title: "Inference"
format: html
editor: visual
---

## Bayesian inference

This chapter contains code to help set up phylogenetic inference pipelines.

## Revbayes

All revbayes scripts can also be downloaded: [fbd_inference_base.rev](fbd_inference_base.rev){download="bd_inference_base.rev"} [model_clock.rev](model_clock.rev){download="model_clock.rev"} [model_morpho.rev](model_morpho.rev){download="model_morpho.rev"} [model_fbd_base.rev](model_fbd_base.rev){download="model_fbd_base.rev"}
[model_fbd_skyline_sampling.rev](model_fbd_skyline_sampling.rev){download="model_fbd_skyline_sampling.rev"} [fbd_inference_skyline.rev](fbd_inference_skyline.rev){download="fbd_inference_skyline.rev"}

### Standard FBD inference

This as a simple setup for a constant rate FBD inference. For more complex cases (e.g., age uncertainty, total evidence etc.) see [this tutorial](https://revbayes.github.io/tutorials/fbd/fbd_specimen.html).

We use this basic setup, available in the file `fbd_inference_base.rev`:

```{r}
#| eval: false
char_mat <- "char_mat.nex"
taxon_ages <- "ages.tsv"
#init_tree_name <- "INSERT_HERE" - not used, if you want to initialize the starting tree
outfile <- "trace"
treefile <- "trees"
outfile_name <- outfile + ".log"
treefile_name <- treefile + ".log"

# import data
morph <- readDiscreteCharacterData(char_mat)
taxa <- readTaxonData(taxon_ages)

# true model parameters to specify initial state & priors if needed
rho <- 1.0
true_origination_rate <- 2
true_extinction_rate <- 0.3
true_sampling_rate <- 10
true_origin_time <- 2
true_clock_rate = 0.5

# initialize moves
moves = VectorMoves()

# load tree model
source("model_fbd_const.rev")
# load clock model
source("model_clock.rev")
# load model of morphological evolution
source("model_morpho.rev")

phy_morpho ~ dnPhyloCTMC(tree = fbd_tree, branchRates = branch_rates_morpho, Q = Q_morpho, type = "Standard", coding = "all")
phy_morpho.clamp(morph)

my_model = model(phy_morpho)
#my_model.graph("fbd_base.dot") generate model graph


# initialize monitors
monitors = VectorMonitors()
# write state to tracefile
monitors.append( mnModel(filename=outfile_name, printgen=10) )
# write tree to tree file
monitors.append(mnFile(filename = treefile_name, printgen = 10, fbd_tree))
# output to the screen
monitors.append(mnScreen(printgen=100, origin_time, speciation_rate, extinction_rate) )

# initialize mcmc
mymcmc = mcmc(my_model, monitors, moves, nruns = 2, ntries = 10000)
mymcmc.burnin(generations=2000,tuningInterval=1500, verbose = TRUE)
mymcmc.operatorSummary()
mymcmc.run(generations = 10000)
mymcmc.operatorSummary()

q()
```

Note that the "true" model parameters are provided. This is to accelerate convergence by defining appropriate priors and set the initial state of the analysis. Empirically, this is would never be the case.

The base analysis calls the three key model components: clock, morphological model, and tree model.

The clock model is specified in `model_clock.rev`:

```{r}
#| eval: false
# strict morphological clock
branch_rates_morpho ~ dnExponential(lambda = 1/true_clock_rate) 
branch_rates_morpho.setValue(true_clock_rate)
moves.append( mvScale(branch_rates_morpho, weight = 5.0, lambda = 0.1))
moves.append( mvScale(branch_rates_morpho, weight = 5.0, lambda = 1.0))
moves.append( mvScale(branch_rates_morpho, weight = 5.0, lambda = 10.0))
```

The morphological model is specified in `model_morpho.rev`:

```{r}
#| eval: false
# model of morphological evolution
# Mk2 model  - no ascertaintment correction
Q_morpho := fnJC(2)
```

And last, the tree model is specified in `model_fbd_tree.rev`:

```{r}
#| eval: false
speciation_rate ~ dnExponential(lambda = 1/true_origination_rate) # set to true mean
moves.append( mvScale(speciation_rate, weight=5.0, lambda = 0.1) )
moves.append( mvScale(speciation_rate, weight=5.0, lambda = 1.0) )
moves.append( mvScale(speciation_rate, weight=5.0, lambda = 5.0) )
speciation_rate.setValue(true_origination_rate)

extinction_rate ~ dnExponential(lambda = 1/true_extinction_rate) # set to true mean
moves.append( mvScale(extinction_rate, weight=5.0, lambda = 0.1) )
moves.append( mvScale(extinction_rate, weight=5.0, lambda = 1.0) )
moves.append( mvScale(extinction_rate, weight=5.0, lambda = 5.0) )
extinction_rate.setValue(true_extinction_rate)

origin_time ~ dnExponential(lambda = 1.0)
moves.append( mvScale(origin_time, weight=5.0, lambda = 0.1) )
moves.append( mvScale(origin_time, weight=5.0, lambda = 1.0) )
moves.append( mvScale(origin_time, weight=5.0, lambda = 5.0) )
origin_time.setValue(true_origin_time)

psi ~ dnExponential(lambda = 1/true_sampling_rate)
psi.setValue(true_sampling_rate)
moves.append( mvScale( psi, weight = 5.0, lambda = 0.1) )
moves.append( mvScale( psi, weight = 5.0, lambda = 1.0) )
moves.append( mvScale( psi, weight = 5.0, lambda = 10.0) )


fbd_tree ~ dnFBDP(origin=origin_time, lambda=speciation_rate, mu=extinction_rate, psi=psi, rho=rho, taxa=taxa)

moves.append( mvFNPR(fbd_tree, weight=20.0) )
moves.append( mvGPR(fbd_tree, weight = 10.0))
moves.append( mvCollapseExpandFossilBranch(fbd_tree, origin_time, weight=50.0) )
moves.append( mvNodeTimeSlideUniform(fbd_tree, weight=10.0) )
moves.append( mvNodeTimeScale(fbd_tree, weight = 20.0))
moves.append( mvRootTimeSlideUniform(fbd_tree, origin_time, weight=50.0) )
moves.append( mvSubtreeScale(fbd_tree, weight = 10.0))
moves.append( mvTreeScale(fbd_tree, origin_time, weight = 10.0))
```

With all these parts in place, you can run the analysis

```{r}
#| eval: false
rb fbd_inference.rev
```

in your terminal. This might take a while, and there are useful tools such as `nohup` (for Linux) that allow you to run these analyses in the background:

```{r}
#| eval: false
nohup rb fbd_inference.rev > out.log 2>&1 &
```


### Skyline FBD analysis

For the FBD skyline sampling analysis that formalizes stratigraphic knowledge as priors we use a very similar base setup, available in `fbd_inference_skyline.rev`:

```{r}
#| eval: false
char_mat <- "char_mat.nex"
taxon_ages <- "ages.tsv"
#init_tree_name <- "INSERT_HERE" - not used, if you want to initialize the starting tree
outfile <- "trace_skyline"
treefile <- "trees_skyline"
outfile_name <- outfile + ".log"
treefile_name <- treefile + ".log"

# import data
morph <- readDiscreteCharacterData(char_mat)
taxa <- readTaxonData(taxon_ages)

# true model parameters to specify initial state & priors if needed
rho <- 1.0
true_origination_rate <- 2
true_extinction_rate <- 0.3
true_origin_time <- 2
true_clock_rate = 0.5

# initialize moves
moves = VectorMoves()

# load tree model
source("model_fbd_skyline_sampling.rev")
# load clock model
source("model_clock.rev")
# load model of morphological evolution
source("model_morpho.rev")

phy_morpho ~ dnPhyloCTMC(tree = fbd_tree, branchRates = branch_rates_morpho, Q = Q_morpho, type = "Standard", coding = "all")
phy_morpho.clamp(morph)

my_model = model(phy_morpho)
#my_model.graph("fbd_base.dot") generate model graph


# initialize monitors
monitors = VectorMonitors()
# write state to tracefile
monitors.append( mnModel(filename=outfile_name, printgen=10) )
# write tree to tree file
monitors.append(mnFile(filename = treefile_name, printgen = 10, fbd_tree))
# output to the screen
monitors.append(mnScreen(printgen=100, origin_time, speciation_rate, extinction_rate, psi) )

# initialize mcmc
mymcmc = mcmc(my_model, monitors, moves, nruns = 2, ntries = 10000)
mymcmc.burnin(generations=2000,tuningInterval=1500, verbose = TRUE)
mymcmc.operatorSummary()
mymcmc.run(generations = 10000)
mymcmc.operatorSummary()

q()
```

The tree model is slightly different because of the skyline analysis. The contents of `model_fbd_skyline_sampling.rev` are

```{r}
#| eval: false
speciation_rate ~ dnExponential(lambda = 1/true_origination_rate) # set to true mean
moves.append( mvScale(speciation_rate, weight=5.0, lambda = 0.1) )
moves.append( mvScale(speciation_rate, weight=5.0, lambda = 1.0) )
moves.append( mvScale(speciation_rate, weight=5.0, lambda = 5.0) )
speciation_rate.setValue(true_origination_rate)

extinction_rate ~ dnExponential(lambda = 1/true_extinction_rate) # set to true mean
moves.append( mvScale(extinction_rate, weight=5.0, lambda = 0.1) )
moves.append( mvScale(extinction_rate, weight=5.0, lambda = 1.0) )
moves.append( mvScale(extinction_rate, weight=5.0, lambda = 5.0) )
extinction_rate.setValue(true_extinction_rate)

origin_time ~ dnExponential(lambda = 1.0)
moves.append( mvScale(origin_time, weight=5.0, lambda = 0.1) )
moves.append( mvScale(origin_time, weight=5.0, lambda = 1.0) )
moves.append( mvScale(origin_time, weight=5.0, lambda = 5.0) )
origin_time.setValue(true_origin_time)

# skyline breakpoints from the example age-depth model
# focusing on gaps longer than 0.1 Myr
timeline <- v(0.0978, 0.2250, 0.2772, 0.4090, 0.4622, 0.5702)

psi[1] ~ dnExponential(lambda = 0.2)
psi[2] <- 0.00001 # fix to small numeric value to formalize that no fossils are sampled in the gap
psi[3] ~ dnExponential(lambda = 0.2)
psi[4] <- 0.00001
psi[5] ~ dnExponential(lambda = 0.2)
psi[6] <- 0.00001
psi[7] ~ dnExponential(lambda = 0.2)

moves.append( mvScale( psi[1], weight = 5.0, lambda = 0.1) )
moves.append( mvScale( psi[1], weight = 5.0, lambda = 1.0) )
moves.append( mvScale( psi[1], weight = 5.0, lambda = 10.0) )
moves.append( mvScale( psi[3], weight = 5.0, lambda = 0.1) )
moves.append( mvScale( psi[3], weight = 5.0, lambda = 1.0) )
moves.append( mvScale( psi[3], weight = 5.0, lambda = 10.0) )
moves.append( mvScale( psi[5], weight = 5.0, lambda = 0.1) )
moves.append( mvScale( psi[5], weight = 5.0, lambda = 1.0) )
moves.append( mvScale( psi[5], weight = 5.0, lambda = 10.0) )
moves.append( mvScale( psi[7], weight = 5.0, lambda = 0.1) )
moves.append( mvScale( psi[7], weight = 5.0, lambda = 1.0) )
moves.append( mvScale( psi[7], weight = 5.0, lambda = 10.0) )

fbd_tree ~ dnFBDP(origin=origin_time, lambda=speciation_rate, mu=extinction_rate, psi=psi, rho=rho, taxa=taxa, timeline = timeline)

#fbd_tree.setValue(init_tree)
moves.append( mvFNPR(fbd_tree, weight=20.0) )
moves.append( mvGPR(fbd_tree, weight = 10.0))
moves.append( mvCollapseExpandFossilBranch(fbd_tree, origin_time, weight=500.0) )
moves.append( mvNodeTimeSlideUniform(fbd_tree, weight=10.0) )
moves.append( mvNodeTimeScale(fbd_tree, weight = 20.0))
moves.append( mvRootTimeSlideUniform(fbd_tree, origin_time, weight=0.0) )
moves.append( mvSubtreeScale(fbd_tree, weight = 100.0))
moves.append( mvTreeScale(fbd_tree, origin_time, weight = 100.0))
```

