---
title: "Lobsters of the deep"
format: html
editor: visual
---

We found a few bugs - towards the bottom to do with reading in a tree with real data in including tip labels and branch lengths

Polychelidans are deep sea lobsters. In the Mesozoic they lived in shallow water. We are going to try and simulate a dataset where the taxa we are recovering originally live in shallow water and then move into deep water. Can we recover the original tree?

QUESTION: We would ideally like to be able to moderate different niches for different parts of the tree **Libraries**

```{r}
setwd("C:/Users/deardenr/OneDrive - University of Birmingham/Documents/Lobsters_of_the_deep")
library(TreeSim)
library(ape)
library(MorphSim)
```

**Environment** We are reusing the enviroment we originally created in Julia because we don't want ot run Julia again.

```{r}
  adm_location = "C:/Users/deardenr/stratworkshop/Lorentz-Workshop/book/data/output/columns_adm.csv" # adjust if you have a different output location
data_adm = read.csv(adm_location) # read age-depth model data
key = "mid"
t =  data_adm[,which(grepl("Myr",names(data_adm)))] # time step
h = data_adm[,which(grepl(key,names(data_adm)))] # heights at the location with key in the name
# define age-depth model
adm = admtools::tp_to_adm(t = t,
                          h = h,
                          T_unit = "Myr",
                          L_unit = "m")

plot(adm,
     lwd_destr = 1, # line width destructive interval (hiatus)
     lwd_acc = 2, # line width accumulative interval
     col_destr = NULL,
     col_acc = "black")
admtools::T_axis_lab()
admtools::L_axis_lab()
title(main = "Age-depth model")
```

**Tree** We are using Myrthe's Polychelid consensus tree and treating this as the "true" tree.

```{r}
 tree <- read.nexus("Lobsters_of_the_deep/TheTrueTree.tree")
tree$edge.length<-
   tree$edge.length/max(nodeHeights(tree)[,2])*1
plot(tree)
```

**Sampling fossils on the tree**

```{r}
seed = 42
lambda = 3 # origination rate [Myr^-1]
mu = 0.7 # extinction rate [Myr^-1]
psi = 20 # fossil sampling rate [fossils per Myr per lineage]
set.seed(seed)
# timescale of observation - depends on duration of your forward simulation
t_max = admtools::max_time(adm)

fossils = FossilSim::sim.fossils.poisson(rate = psi,
                                         tree = tree,
                                         root.edge = FALSE)


plot(fossils, 
     tree = tree,
     show.ranges=TRUE,
     main = "FBD tree")
```

**Ecological and taphonomic effects**

```{r}
wd_location = "C:/Users/deardenr/stratworkshop/Lorentz-Workshop/book/data/output/columns_wd.csv" # adjust if you have a different output location
data_wd = read.csv(wd_location) # read age-depth model data
# select water depth column that contains the key
wd = data_wd[,which(grepl(key,names(data_wd)))]
plot(x = t,
     y = wd,
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Water depth [m]",
     main = "Water depth")
```

LOBSTERS OF THE SHALLOWS These lobsters like living in shallow water

```{r}
wd_opt = 5 # pref. water depth 5 m
wd_tolerance = 5 # water depth tolerance 10 m (sd)
cutoff_val = 0  # does not live above water 
niche_def_s = StratPal::snd_niche(opt = wd_opt, 
                                tol = wd_tolerance, 
                                cutoff_val = cutoff_val) 
wd_vals_s = seq(from = min(wd),
              to = max(wd),
              length.out = 100)
plot(x = wd_vals_s,
     y = niche_def_s(wd_vals_s),
     type = "l",
     xlab = "Water depth [m]",
     ylab = "Sampling probability",
     main = "Niche definition")
```

LOBSTERS OF THE DEEP

```{r}
wd_opt = 17 # pref. water depth 5 m
wd_tolerance = 3 # water depth tolerance 10 m (sd)
cutoff_val = 0  # does not live above water 
niche_def_d = StratPal::snd_niche(opt = wd_opt, 
                                tol = wd_tolerance, 
                                cutoff_val = cutoff_val) 
wd_vals_d = seq(from = min(wd),
              to = max(wd),
              length.out = 100)
plot(x = wd_vals_d,
     y = niche_def_d(wd_vals_d),
     type = "l",
     xlab = "Water depth [m]",
     ylab = "Sampling probability",
     main = "Niche definition")
```

Define the gradient function

```{r}
# define function that specifies water depth as a function of time
gradient = approxfun(x = t,
                     y = wd)
plot(x = t,
     y = gradient(t),
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Water depth [m]",
     main = "Water depth as gradient")
```

**Sampling compared** Shallow lobs

```{r}
plot(x = t,
     y = t |> gradient() |> niche_def_s(),
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Sampling probability",
     main = "Sampling probability of shallow lobs, time domain")
```

Deep lobs

```{r}
plot(x = t,
     y = t |> gradient() |> niche_def_d(),
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Sampling probability",
     main = "Sampling probability of deep lobs, time domain")
```

**plotting tree baseline**

```{r}
# Baseline plot
plot(fossils, 
     tree = tree,
     main = "Full fossil record without ecological effects",
     rho = 0)
```

**plotting tree shallow**

```{r}
fossils_after_ecol_s = fossils |>
  admtools::rev_dir(ref = t_max) |>
  StratPal::apply_niche(niche_def = niche_def_s, gc = gradient) |>
  admtools::rev_dir(ref = t_max)

rho=1
fossils_after_ecol_s_ext = FossilSim::sim.extant.samples(fossils = fossils_after_ecol_s,
                                         tree = tree,
                                         rho = rho)


plot(fossils_after_ecol_s_ext, 
     tree = tree,
     main = "FBD tree after ecological effects shallow",
     show.tip.label = TRUE,
     cex=0.5)
```

```{r}
fossils_after_ecol_d = fossils |>
  admtools::rev_dir(ref = t_max) |>
  StratPal::apply_niche(niche_def = niche_def_d, gc = gradient) |>
  admtools::rev_dir(ref = t_max)



rho=1
fossils_after_ecol_d_ext = FossilSim::sim.extant.samples(fossils = fossils_after_ecol_d,
                                         tree = tree,
                                         rho = rho)


plot(fossils_after_ecol_d_ext, 
     tree = tree,
     main = "FBD tree after ecological effects deep",
     show.tip.label = TRUE,
     cex=0.5)
```

**BUG** **Sampled trees**

```{r}
plot(fossils_after_ecol_s_ext,
     tree = tree,
     extant.col = "red",
     reconstructed = FALSE,
     main = "Sampled tree shallow")

plot(fossils_after_ecol_d_ext,
     tree = tree,
     extant.col = "red",
     reconstructed = TRUE,
     main = "Sampled tree deep")

library(MorphSim)
plot(data = fossils_after_ecol_s_ext, trait = 1, timetree = T)
```

**Shallow lob matrix**

```{r}
# simple Mk2 model
n_states = 2
br_rate = 0.5
trait_num = 10
trait_to_display = 2
morpho_mk_s = MorphSim::sim.morpho(k = n_states,
                        time.tree = tree,
                        br.rates = br_rate,
                        trait.num = trait_num, 
                        variable = FALSE,
                        fossil = fossils_after_ecol_s_ext)
MorphSim::plotMorphoGrid(morpho_mk_s)
#Bug here when plotting?
plot(morpho_mk_s,
     trait = trait_to_display)
title(main = paste0("Trait no. ", trait_to_display))
```

**Deep lob matrix**

```{r}
# simple Mk2 model
n_states = 2
br_rate = 0.5
trait_num = 10
trait_to_display = 2
morpho_mk_d = MorphSim::sim.morpho(k = n_states,
                        time.tree = tree,
                        br.rates = br_rate,
                        trait.num = trait_num, 
                        variable = FALSE,
                        fossil = fossils_after_ecol_d_ext)
MorphSim::plotMorphoGrid(morpho_mk_d)
#Bug here when plotting?
plot(morpho_mk_d,
     trait = trait_to_display)
title(main = paste0("Trait no. ", trait_to_display))
```

```{r}
# export character matrix
MorphSim::write.morpho(data = morpho_mk_d,
                       file = "char_mat_temp.nex",
                       type = "matrix",
                       all = FALSE,
                       reconstructed = TRUE)
# export taxon ages
MorphSim::write.morpho(data = morpho_mk,
                       file = "ages_temp.tsv",
                       type = "ages",
                       all = TRUE,
                       reconstructed = TRUE)

# export the true tree 
MorphSim::write.morpho(data = morpho_mk,
                       file = "true_tree.tre",
                       type = "tree",
                       all = TRUE,
                       reconstructed = TRUE)
```

Nothign is working and we can't wxport the matrix so we're going to make some specialised deep sea lobster datasets

**Deep sea lobster**

```{r}
# simple Mk2 model
n_states = 2
br_rate = 0.5
trait_num = 10
trait_to_display = 2
morpho_mk_d = MorphSim::sim.morpho(k = n_states,
                        time.tree = tree,
                        br.rates = br_rate,
                        trait.num = trait_num, 
                        variable = FALSE,
                        fossil = fossils_after_ecol_d_ext)
MorphSim::plotMorphoGrid(morpho_mk_d)
#Bug here when plotting?
plot(morpho_mk_d,
     trait = trait_to_display)
title(main = paste0("Trait no. ", trait_to_display))
```
