---
title: "Phylogeny and stratigraphy"
format: html
editor: visual
---

## Introduction

This tutorial combines everything we have learned so far by combining stratigraphic paleobiology and phylogenetics into one forward simulation.

## Age-depth model and environmental gradient

Let's begin by loading the age-depth model data. For simplicity, we focus on the mid-platform section only

```{r}
adm_location = "../data/output/columns_adm.csv" # adjust if you have a different output location
data_adm = read.csv(adm_location) # read age-depth model data
key = "mid"

t =  data_adm[,names(data_adm)[2]] # time step
# select height column that contains the key
h = data_adm[,which(grepl(key,names(data_adm)))]
# define age-depth model
adm = admtools::tp_to_adm(t = t,
                          h = h,
                          T_unit = "Myr",
                          L_unit = "m")
plot(adm,
     lwd_destr = 1, # line width destructive interval (hiatus)
     lwd_acc = 2, # line width accumulative interval
     col_destr = NULL,
     col_acc = "black")
admtools::T_axis_lab() # add time axis label
admtools::L_axis_lab() # add length axis label (stratigraphic domain)
```

For an environemental gradient, we will focus on water depth:

```{r}
wd_location = "../data/output/columns_wd.csv" # adjust if you have a different output location
data_wd = read.csv(wd_location) # read age-depth model data
key = "mid"
t =  data_wd[,names(data_wd)[2]] # time step
# select height column that contains the key
wd = data_wd[,which(grepl(key,names(data_wd)))]

plot(x = t,
     y = wd,
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Water depth [m]",
     main = "Water depth")
```

## Constants

For convenience, let's introduce some constants:

```{r}
seed = 42
t_max = admtools::max_time(adm) # duration of the observed time interval
h_max = admtools::max_height(adm)
lambda = 2
mu = 0.3
psi = 70
```

## Simulating the tree and fossils

```{r}
set.seed(seed)
tree = TreeSim::sim.bd.age(age = t_max,
                              numbsim = 1,
                              lambda = lambda,
                              mu = mu,
                              frac = 1,
                              mrca = FALSE,
                              complete = TRUE)[[1]]
plot(tree)
```

Now let's simulate some fossils along that tree:

```{r}
fossils = FossilSim::sim.fossils.poisson(rate = psi,
                                         tree = tree)
plot(fossils, tree)
```

this is all in the time domain.

## Ecological and taphonomic assumptions

We focus on ecological preferences here, taphonomy works mutatis mutandis.

```{r}
gc = approxfun(x = t, # gradient change with time
               y = wd)
niche_def = StratPal::snd_niche(opt = 5, # preferred water depth 5 m
                      tol = 10, # tolerance to fluctuations 10 m (sd)
                      cutoff_val = 0) # no preservation on land
wd_range = range(wd)
wd_vals = seq(min(wd), max(wd), by = 0.1)
plot(wd_vals,
     niche_def(wd_vals),
     type = "l",
     xlab = "Water depth [m]",
     ylab = "Preservation probability",
     main = "Ecological niche")
```

Now we can determine the preservation probability in the time domain:

```{r}
plot(t, t |> gc() |> niche_def(),
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Collection probability")
```

Note that here the collection probability drops to 0 during gaps NOT because of erosion and gaps, but because the modeled taxa are strictly marine

## Observed tree and fossils in the stratigraphic domain

Let's see what we can observe in the stratigraphic domain:

```{r}
tree_strat = tree |> 
  admtools::time_to_strat(adm) |>
  plot()
axis(side = 1, labels = TRUE)
mtext("Height", side = 2, line = 1.4)
```

You can nicely see how branching events that coincide with gaps are expressed as polytomies in the stratigraphic domain.

Now let's add some fossils:

Note the rev_dir to account for the fact that age-depth models are forward in time and trees use age:

```{r, eval=FALSE}
fossils_strat = fossils |> 
  admtools::rev_dir(ref = t_max) |>
  admtools::time_to_strat(adm) |>
  admtools::rev_dir(ref = h_max) |>
  plot(tree_strat)
axis(side = 1, labels = TRUE)
mtext("Height", side = 2, line = 1.4)
```

## Modifying the fossil record

All stratigraphic paleobiology pipelines as described before work with fossil objects.

```{r}
f = StratPal::strat_filter(adm)

fossils_filtered = StratPal::apply_taphonomy(fossils, identity, f)
print(fossils_filtered)

fossils |>
  admtools::rev_dir(ref = t_max) |>
  StratPal::apply_taphonomy(identity, f) |>
  StratPal::apply_niche(niche_def, gc) |>
  admtools::rev_dir(ref = t_max) |>
  plot(tree)
```

```{r}
morpho_mk <- MorphSim::sim.morpho(k = 2, time.tree = tree, br.rates = 0.1, trait.num = 15,  fossil = fossils)
MorphSim::plotMorphoGrid(morpho_mk)
MorphSim::write.morpho(morpho_mk, file = "hello.nex", reconstructed = FALSE)
```

```{r}
plot(x = morpho_mk, 
     trait= 2, 
     timetree = T, 
     show.fossil = T,
     root.edge = T, 
     reconstructed = T)
```

```{r}

```
