---
title: "Phylogeny and stratigraphy"
format: html
editor: visual
---

## Introduction

This chapter combines everything we have learned so far by combining stratigraphic paleobiology and phylogenetics into one forward simulation. At the end of this chapter, you will have built a complete simulation pipeline consisting of

-   A stratigraphic forward model (CarboKitten) that simulates age-depth models and environmental gradients such as water depth and facies

-   A FBD tree that describes the evolutionary relationship between taxa

-   Fossils sampled along the tree according to an originally constant sampling rate, which is modified by gaps in the stratigraphic record and ecological as well as taphonomic effects

-   A character matrix for the fossils evolving according to a specified model of evolution and potentially missing characters due to differential preservation

-   Fossil ages that are potentially offset due to age misassignment or uncertain

With all these pieces in place, we can then address two key questions in the next chapter:

1.  When we run inference on the simulated data (fossil ages, character matrices), can we recover the "true" evolutionary history (e.g., tree topology, model parameters)? More speculatively, can we potentially recover stratigraphic information such as sediment accumulation rates and gap duration from phylogenetic inference?
2.  Having build the forward simulation pipeline, can we use our knowledge of stratigraphy and the sampling process as sources of information in the inference? And if yes, does it improve the inference?

Your **task** for this chapter is to modify the provided pipeline according to your work in the last 3 days and simulate your very own fossil record. At the end you should have two components: (1) simulated biased data that you would usually use for inference, specifically a character matrix and fossil ages and (2) the biological and stratigraphic "truth" to compare inference results to. Specifically, this should contain the true sampled tree, age-depth relationship, clock rate, and model of morphological evolution.

## True and assumed age-depth model

Let's begin by loading the age-depth model data. For simplicity, we focus on the mid-platform section only

```{r}
#| layout-ncol: 2
adm_location = "../data/output/columns_adm.csv" 
data_adm = read.csv(adm_location) # read age-depth model data
key = "mid"

t =  data_adm[,which(grepl("Myr", names(data_adm)))] # time step
# select height column that contains the key
h = data_adm[,which(grepl(key,names(data_adm)))]
# define age-depth model
adm = admtools::tp_to_adm(t = t,
                          h = h,
                          T_unit = "Myr",
                          L_unit = "m")

h_max = admtools::max_height(adm)
h_min = admtools::min_height(adm)
t_max = admtools::max_time(adm)
t_min = admtools::min_height(adm)

adm_assumed = admtools::tp_to_adm(t = c(t_min, t_max),
                                  h = c(h_min, h_max),
                                  T_unit = "Myr",
                                  L_unit = "m")
plot(adm,
     lwd_destr = 1, # line width destructive interval (hiatus)
     lwd_acc = 2, # line width accumulative interval
     col_destr = NULL,
     col_acc = "black")
admtools::T_axis_lab() # add time axis label
admtools::L_axis_lab() # add length axis label (stratigraphic domain)
title(main = "True age-depth model")

plot(adm_assumed,
     lwd_destr = 1, # line width destructive interval (hiatus)
     lwd_acc = 2, # line width accumulative interval
     col_destr = NULL,
     col_acc = "black")
admtools::T_axis_lab() # add time axis label
admtools::L_axis_lab() # add length axis label (stratigraphic domain)
title(main = "Assumed age-depth model")
```

## Taphonomy and ecology

Again, we use water depth as ecological gradient:

```{r}
wd_opt = 5
wd_tol = 10
cutoff_val = 0
#| layout-ncol: 2
wd_location = "../data/output/columns_wd.csv" # adjust if you have a different output location
data_wd = read.csv(wd_location) # read age-depth model data
t =  data_wd[,names(data_wd)[2]] # time step
# select height column that contains the key
wd = data_wd[,which(grepl(key,names(data_wd)))]

gradient = approxfun(x = t,
                     y = wd)
plot(x = t,
     y = gradient(t),
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Water depth [m]",
     main = "Water depth as gradient")

niche_def = StratPal::snd_niche(opt = wd_opt, # pref. water depth 5 m
                                tol = wd_tol, # water depth tolerance 10 m (sd)
                                cutoff_val = cutoff_val) # does not live above water 
wd_vals = seq(from = min(wd),
              to = max(wd),
              length.out = 100)
plot(x = wd_vals,
     y = niche_def(wd_vals),
     type = "l",
     xlab = "Water depth [m]",
     ylab = "Sampling probability",
     main = "Niche definition")
```

This results in the following sampling probability:

```{r}
plot(x = t,
     y = t |> gradient() |> niche_def(),
     type = "l",
     xlab = "Time",
     ylab = "Sampling probability",
     main = "Sampling probability of fossils")
```

For the taphonomy, we again remove all fossils coinciding with gaps:

```{r}
gap_remove = StratPal::strat_filter(adm)
plot(t, gap_remove(t),
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Sampling probability")
```

## FBD simulation

Next, simulate the tree and the fossils along it:

```{r}
seed = 42
set.seed(seed)
lambda = 2
mu = 0.3
psi = 70
tree = TreeSim::sim.bd.age(age = t_max,
                              numbsim = 1,
                              lambda = lambda,
                              mu = mu,
                              frac = 1,
                              mrca = FALSE,
                              complete = TRUE)[[1]]
fossils = FossilSim::sim.fossils.poisson(rate = psi,
                                         tree = tree)
plot(fossils, tree)
```

## Taphonomic and ecological effects on fossils

Now we can apply our ecological and taphonomic filters to the fossil object:

```{r}
fossils_pres = fossils |>
  admtools::rev_dir(ref = t_max) |>
  StratPal::apply_niche(niche_def = niche_def,
                        gc = gradient) |>
  StratPal::apply_taphonomy(pres_potential = gap_remove,
                            ctc = identity) |>
  admtools::rev_dir(ref = t_max)
plot(fossils_pres, tree)
```

Note that we can combine the original fossil sampling rate `psi` with the ecological and taphonomic effects to determine the continuously varying sampling rate in time:

```{r}
total_sampling_rate = function(x){
  y = psi * niche_def(gradient(x)) * gap_remove(x)
  return(y)
}
# use lower temporal resolution
t_low_res = seq(t_min, t_max, length.out = 100)
plot(x = t_low_res,
     y = total_sampling_rate(t_low_res),
     type = "l",
     xlab = "Time",
     ylab = "Effective sampling rate")
```

## Simulate morphological characters

Now we can simulate morphological characters along the tree and the preserved fossils:

```{r}
n_states = 2
br_rate = 0.1
trait_num = 15
morpho_mkv = MorphSim::sim.morpho(k = n_states,
                        time.tree = tree,
                        br.rates = br_rate,
                        trait.num = trait_num, 
                        variable = FALSE,
                        fossil = fossils_pres)
```

## Modify ages

Now we can modify the ages stored in the fossil object:

```{r}
fossils_age_misassigned = morpho_mkv$fossil |> 
  admtools::rev_dir(ref = t_max) |>
  admtools::time_to_strat(adm) |> 
  admtools::strat_to_time(adm_assumed) |>
  admtools::rev_dir(ref = t_max)
# replace correct ages with misassigned ages
# for easier export via MorphSim::write.morpho
morpho_mkv$fossil = fossils_age_misassigned
```

Do you know why this is only done now and not in one of the earlier steps that modify the fossils?

## Export data

Finally, we can export the tree, character matrix, and fossil ages
