---
title: "MorphSim Tutorial"
---

## Simulating phylogenetic trees in R

Before we can simulate morphological data we need a phylogenetic tree along which to simulate our data. In R, trees can be simulated using a number of different packages, most commonly `ape` and `TreeSim`. Here we explain how to generate a birth-death time tree using `TreeSim`.

`TreeSim` (Stadler 2011) simulates trees under a range of birth-death processes, where new lineages are added at rate `lambda` (speciation) and removed at rate `mu` (extinction). The function we'll use, `sim.bd.taxa()`, simulates a tree conditioned on reaching a fixed number of extant tips.

`TreeSim` provides several related functions depending on what you want to condition the simulation on:

| Function | Conditions on |
|------------------------------------|------------------------------------|
| `sim.bd.taxa()` | number of tips |
| `sim.bd.age()` | tree age (time since root) |
| `sim.rateshift.taxa()` | number of tips, with rate shifts / mass extinctions over time |

We'll focus on `sim.bd.taxa()` since it's the simplest entry point, but it's worth knowing the others exist if your own data calls for a different kind of tree.

For reproducibility, we set a random seed before simulating: since the birth-death process is stochastic, this ensures the tree (and everything downstream) is identical every time this document is rendered.

```{r}
library(TreeSim)
library(ape)

set.seed(12)
```

### Generating the tree

We simulate a single tree with 10 tips, speciation rate of 1 and an extinction rate of 0.5. Note that `mu` must be less than `lambda`; if extinction is too close to (or exceeds) speciation, the process may need many attempts to reach the target number of tips, and simulation will be slow or may not converge in reasonable time. Note that the values chosen. for `mu` and `lambda` are arbitrary here should be chosen carefully. Most empirical b/d rates are much lower than this.

```{r}
tree_list <- sim.bd.taxa(
  n        = 10,   # number of extant tips
  numbsim  = 1,    # number of trees to simulate
  lambda   = 1,    # speciation rate
  mu       = 0.5,  # extinction rate
  complete = TRUE # if set to false drop all extinct lineages and keep only extant tips
)

tree <- tree_list[[1]]
```

Note that `sim.bd.taxa()` always returns a **list** of trees, with length equal to `numbsim`. Even when simulating a single tree, we still need to index into the list (`[[1]]`) to extract the `phylo` object itself. This trips a lot of people up the first time, so it's worth calling out explicitly.

We set `complete = TRUE` so we want the whole tree, including lineages that went extinct before the present, rather than one pruned down to extant taxa only. It's important to understand what `n` controls here: `sim.bd.taxa()` only lets you specify the number of extant tips, not the total number of tips in the tree. So with `n = 10` and `complete = TRUE`, the simulation keeps adding and removing lineages until it reaches exactly 10 tips at the present, but every lineage that went extinct along the way is also retained in the output. That means the final tree will generally have more than 10 tips in total once extinct lineages are counted. If you instead want a tree with exactly 10 tips and no extinct lineages at all, set `complete = FALSE`, which drops the extinct lineages and returns only the extant-tip tree.

### Checking the tree

Before moving on to simulating morphology, it's good practice to confirm the tree looks the way we expect. A few quick checks:

```{r}
# total tip count, including BOTH extant and extinct lineages
# (this will be > 10, since n only controls extant tips)
Ntip(tree)
```

```{r}
# this is the check that actually confirms n = 10 was enforced:
# count tips whose distance from the root equals the tree's max depth
# (i.e. tips sampled at the present)
node_depths <- node.depth.edgelength(tree)
tip_depths  <- node_depths[1:Ntip(tree)]
sum(abs(tip_depths - max(tip_depths)) < 1e-8)
```

```{r}
# with complete = TRUE, the tree should generally NOT be ultrametric,
# since extinct tips end before the present
is.ultrametric(tree)
```

```{r}
# TreeSim always produces a strictly bifurcating tree so this
# should be TRUE regardless of complete = TRUE/FALSE
is.binary(tree)
```

### Visualizing the tree

A quick plot lets us see the extinct tips (branches ending before the present) alongside the extant ones:

```{r}

plot(tree, cex = 0.8, no.margin = TRUE)

```

With a validated `phylo` object, extant and extinct tips both retained, we're ready to simulate morphological character data along its branches using `MorphSim`.

## Simulating Morphological Data

Now that we have a validated tree (extant tips plus extinct lineages, from `TreeSim`. We will simulate fossil sampling using the FBD model tomorrow!), we can simulate morphological character data along it using `MorphSim`. We'll build up the model one assumption at a time, starting from the simplest possible case and progressively relaxing it, the same tree is used throughout, so any differences in the resulting data are purely down to the model of morphological evolution, not the tree.

```{r}
library(MorphSim)
```

### The baseline: Mk

The simplest model for discrete morphological data is the **Mk model** (Lewis 2001), a direct extension of the Jukes-Cantor idea to morphology: `k` possible states per character, one shared substitution rate, and no direction favoured over any other. It's a strong assumption. Unlike a nucleotide, state "1" in one character has no inherent relationship to state "1" in another, but it's the natural starting point.

```{r}
morpho_mk <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15)
plotMorphoGrid(morpho_mk)
```

Take a look at the grid, you'll likely spot at least one invariant column (a character that never changes across any tip). Real morphological datasets essentially never look like this, for a specific reason we'll get to next. Note that by setting `k = 3` we are simulating traits 0,1,2.

### MkV: conditioning on variability

Empirical matrices are curated by systematists, and a character that doesn't vary across the study group is rarely recorded, invariant characters simply aren't often scored in the first place. If we simulated data and then deleted invariant columns after the fact, we'd bias our branch length and rate estimates, since we'd have thrown away information about how rare change actually is. MkV builds the exclusion directly into the model, so the simulation (and the corresponding likelihood, if you're fitting a model back to this data) correctly accounts for the fact that we conditioned on variability from the start.

```{r}
morpho_mkv <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1,
                          trait.num = 15, variable = TRUE)

```


### Making sure the state space is fully sampled

Even with `k = 3` allowed, a finite simulation on a finite tree might just never sample one of those states, especially with a small tree, low rates, or few characters. If you're using simulated data to test a 3-state model but only two states ever show up at the tips, you've unintentionally built a test case for a different model than you think you have. `full.states` forces the simulation to keep resampling until every state you specified actually for each character trait.

```{r}
morpho_full <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1,
                           trait.num = 15, variable = TRUE, full.states = TRUE)
```

### Filtering for parsimony-informative characters

A second, related issue: real matrices are mostly composed of **parsimony-informative** characters, at least two states, each seen in two or more taxa. A raw simulation will happily generate autapomorphies (a state found in only a single taxon) or characters that are technically variable but still uninformative for resolving relationships. Rather than hand-filtering afterwards, `MorphSim` lets you enforce this directly:

```{r}
# tolerate autapomorphies
morpho_pars_standard <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1,
                                    trait.num = 15, variable = TRUE,
                                    parsimony = "standard")

# every state must occur in 2+ taxa = stricter
morpho_pars_strict <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1,
                                  trait.num = 15, variable = TRUE,
                                  parsimony = "strict")
```

::: callout-note
Both `full.states` and `parsimony` work by rejection sampling: generate, check, discard, and retry. If `k` is set too high, rates too low, or the tree too small, this can take a long time to converge, `MorphSim` caps at 1 million iterations and will tell you to relax a parameter if it can't satisfy your constraints.
:::

### Among-character rate variation (ACRV)

So far every character has shared a single rate, which is biologically unlikely: some traits are under tight functional or developmental constraint and barely change, while others are close to neutral and drift quickly. Among-character rate variation lets each character draw its own rate multiplier, typically from a discretized gamma distribution, the same approach used for among-site rate variation in molecular models.

```{r}

morpho_acrv <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15,
                           variable = TRUE, ACRV = "gamma", alpha.gamma = 1, ACRV.ncats = 4)

plotMorphoGrid(morpho_acrv)
```

Same tree, same branch rates, the only thing that changed is the character-rate distribution. Does this look closer to a matrix you'd recognize from real data?

### Custom Q-matrices: encoding your own assumptions

Everything above still assumes equal rates in every direction between states. That's often *wrong* for a specific character rather than merely simplified. Meristic characters, or anything that must pass through intermediate states, shouldn't be able to jump directly from state 0 to state 2. An **ordered** model encodes exactly that constraint:

```{r}
#| label: sim-ordered
ord_Q <- matrix(c(-0.5, 0.5, 0.0,
                   0.3333333, -0.6666667, 0.3333333,
                   0.0, 0.5, -0.5),
                nrow = 3, byrow = TRUE)

morpho_ordered <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15,
                              variable = TRUE, ACRV = "gamma", alpha.gamma = 1,
                              ACRV.ncats = 4, define.Q = ord_Q)

plot(morpho_ordered, trait = 1, box.cex = 2)
```

`plot.morpho()` lets you follow one character's full transition history along the tree, a useful way to see the ordering constraint actually playing out branch by branch, rather than just inferring it from the tip states.

### Partitioning the matrix

So far every character in a given call to `sim.morpho()` has shared the same number of states and the same model. Real morphological matrices are rarely that homogeneous, a matrix might include binary presence/absence characters, three-state characters describing shape variation, and four-state meristic counts, all in the same dataset. `MorphSim` handles this with the `partition` argument: you specify how many traits belong to each partition, and give `k` as a vector with the number of states for each partition (in the same order).

```{r}
#| label: sim-partitions

part_data <- sim.morpho(
  time.tree  = tree,
  k          = c(2, 3, 4),
  trait.num  = 20,
  partition  = c(10, 5, 5),
  br.rates   = 0.1
)

# model specification per partition
part_data$model$Specified
```

Here, the first 10 traits are binary (`k = 2`), the next 5 have three states, and the last 5 have four states. The values in `partition` must sum to `trait.num`, or `sim.morpho()` will throw an error.

::: callout-note
`partition` only lets you vary the number of states (`k`) across groups of characters within a single call. If you want different partitions to use *different models entirely* for example, some ordered with a custom `Q`, others unordered with gamma rate variation you can't do that in one call. Instead, simulate each partition separately with its own `sim.morpho()` call, then merge them with `combine.morpho()`. Both objects must share the same tree, branch lengths, and (if available) fossil object.
:::

```{r}
#| label: combine-partitions

# partition 1: ordered characters with a custom Q matrix
partition_1 <- sim.morpho(time.tree = tree, k = 3, trait.num = 10,
                           br.rates = 0.1, define.Q = ord_Q)

# partition 2: unordered binary characters with MkV + gamma ACRV
partition_2 <- sim.morpho(time.tree = tree, k = 2, trait.num = 10,
                           br.rates = 0.1, variable = TRUE,
                           ACRV = "gamma", alpha.gamma = 1, ACRV.ncats = 4)

combined <- combine.morpho(partition_1, partition_2)

# 20 traits total, model info from both partitions preserved
length(combined$sequences$tips[[1]])
combined$model$Specified
```

### Missing data as a taphonomic layer

Everything up to this point has been about the *evolutionary* process, how characters change along the tree. Real fossil matrices are also shaped by a second, independent process: **preservation**. Not every character that existed is observable in a fossil, and different anatomical regions have very different preservation potential. This is where stratigraphy and taphonomy come back into the picture: the pattern of missing data in a real matrix isn't random noise, it's a signal about what kind of tissue, environment, and depositional setting produced the fossil record you're working with.

`sim.missing.data()` lets us impose that layer on top of a simulated matrix, replacing observed states with `"?"` under different missingness scenarios.

**By partition**: mirrors the character partitions we just built: some anatomical regions (e.g., robust skeletal elements) preserve far more reliably than others (e.g., soft tissue characters).

```{r}
# hard tissue characters (partition 1): 10% missing
# soft tissue characters (partition 2): 70% missing
missing_part <- sim.missing.data(
  data = combined, method = "partition",
  seq = "tips", probability = c(0.1, 0.7)
)
```

**Extinct taxa only**: leaves extant tips fully observed and only introduces missing data in extinct taxa, reflecting the common (and very real) asymmetry that living specimens can be studied completely while fossils cannot.

```{r}
missing_extinct <- sim.missing.data(
  data = combined , method = "extinct",
  seq = "tips", probability = 0.5
)
```

Other available `method` options worth knowing about: `"random"` (uniform missing data across the whole matrix, useful as a naive baseline), `"rate"` (missing data tied to a character's ACRV rate category, useful for testing whether fast- or slow-evolving characters are preferentially lost), and `"trait"` / `"taxa"` / `"trait_taxa"` for targeting specific characters or specimens directly.


### Exporting matrices, trees, and fossil ages

Once you're happy with a simulated dataset, `write.morpho()` provides a single interface for exporting everything you need to move into downstream phylogenetic software (RevBayes, BEAST2, or similar): the tree, the character matrix, and fossil occurrence ages. Note we haven't simulated the process of incomplete sampling or fossilization yet to, used to generate fossil ages - we'll do that tomorrow. The `type` argument controls what gets written, and `reconstructed` controls whether you export the full simulated tree/matrix or only the reconstructed version (i.e., only the sampled lineages, relevant once fossils are involved, see day 3).

```{r}
# full tree
write.morpho(missing_part, file = "tree.tre", type = "tree",
             reconstructed = FALSE)

# character matrix
write.morpho(missing_part, file = "matrix.nex", type = "matrix", reconstructed = FALSE)

# fossil occurrence ages (compatible with RevBayes)
write.morpho(missing_part, file = "ages.tsv", type = "ages")

# ages with associated uncertainty
write.morpho(missing_part, file = "ages.tsv", type = "ages",
             uncertainty = 2)
```


::: callout-important
## Why simulate at all?

Before starting the next section it is worth thinking about why we need simulation studies. It might seem backwards to spend a workshop generating *fake* data when the goal is to understand *real* fossil datasets. But simulation is one of the few tools we have for testing whether our inference methods actually recover what we think they recover.

With real data, we never know the true tree, the true model, or the true pattern of missing data, we're always inferring all three at once, which makes it hard to tell whether a surprising result reflects real biology or a limitation of the method. With simulated data, we control every one of those pieces. That means we can:

-   **Test methods against a known truth.** Simulate under a model with a known Q-matrix, known rate heterogeneity, and known tree, then run inference on the result. If the method recovers something far from what you simulated, that tells you something important about its limits, not about the fossil record.
-   **Separate process from artifact.** When something looks unusual in a real matrix (e.g., surprisingly low homoplasy, or an odd concentration of missing data), simulation lets you ask: *could this pattern arise just from preservation bias, even under a perfectly ordinary evolutionary process?* If yes, that's a caution against over-interpreting the pattern biologically.
-   **Build intuition for what "normal" looks like.** Once you've seen how an Mk matrix, an MkV matrix, and a partitioned matrix with realistic missing data each look and behave differently, you're much better equipped to look at someone else's real dataset and have an informed opinion about what its statistics are (and aren't) telling you.

This is also exactly why we built the model up one assumption at a time today, rather than jumping straight to something maximally realistic: every mismatch between a simple model and a real dataset is potentially meaningful, but only if you know precisely which assumption you relaxed to close that gap.
:::

### Your turn: biology meets the fossil record

Everything in this section has been two separate layers stacked on top of each other: an **evolutionary model** (partitions, rates, Q-matrices) that describes how characters actually change, and a **preservation model** (missing data) that describes what we might actually observe of that change once fossils are involved. Real morphological matrices are the product of both and it's easy to mistake a pattern caused by preservation for one caused by evolution, or vice versa, if you're not thinking about them as distinct layers or processes.

For this exercise, work in groups and design **one scenario** that combines both layers:

**Step 1: the biology.** Choose a partition structure that reflects a real hypothesis about your organism of interest. For example:

-   A skeletal dataset with slow-evolving, ordered meristic characters (limb segment counts) alongside fast-evolving, unordered characters (colour pattern, if it were preservable).
-   A dataset where one partition is under strong functional constraint (low rate, few states) and another that is closer to neutral (higher rate, more states).

Build this with `sim.morpho()`, using `partition` and `k` (or `combine.morpho()` if your partitions need genuinely different models, not just different state numbers).

**Step 2: the taphonomy** Now decide *why* some of this data would be missing in a real fossil dataset, and choose a `sim.missing.data()` method that reflects that reasoning rather than picking probabilities arbitrarily. Questions to ground this in:

-   Which of your partitions corresponds to tissue that preserves well (bone, teeth) versus poorly (integument, soft anatomy)? → `method = "partition"`
-   Are your fossil taxa systematically less complete than your extant taxa, independent of which characters they are? → `method = "extinct"`
-   Is there a specific, well-known taphonomic bias for your group, e.g., a particular character system that's notoriously hard to score from compressed or fragmentary material? → `method = "trait"` or `"trait_taxa"`


**Step 3: Are there some limitations to what we can simulate?** Think about all the models you have seen so far, are there any processes you think we are not capturing that would be important for modeling morphological data from fossils? What assumptions could we include to improve our modeling of morphological data?

## Breakout group discussions 


### Group 1 : Variation in rates of evolution in crinoid body plan evolution

Authors: Lena, Mark, Joel, Charlotte, Xianyi

Generating a partition that aims to capture variation in regions of the crinoid body:

-   Calyx = More likely to be under major body plan constraints fewer character states, slower rates of state transition, & taphonomically less missing data.

-   Arms = ecological traits expected to undergo more rapid evolution more character states, faster rates of state transition, and taphonomically more likely to have missing data.

```{R}
tree_list <- TreeSim::sim.bd.taxa(
    n        = 10,   # number of extant tips
    numbsim  = 1,    # number of trees to simulate
    lambda   = 1,    # speciation rate
    mu       = 0.5,  # extinction rate
    complete = TRUE # if set to false drop all extinct lineages and keep only extant tips
)

tree <- tree_list[[1]]


#partition 1  = calyx
calyx_partition <- sim.morpho(k = c(2, 3), time.tree = tree, br.rates = 0.1, trait.num = 16, partition = c(8,8), full.states=TRUE)

#k=c(2,3) is 2 sets of chars, one with 2 states and one with 3 char states
#trait.num=c(8,8) is two sets of 8 chars each
#br.rates = pretty slow
#define.Q = ord_Q is an ordered modeled of character change
#full.states=TRUE means all char. states will be included in the simulated trait data



#Partition 2 = arms
arms_partition <- sim.morpho(k=c(4,5), time.tree=tree, br.rates=0.3, trait.num=16, partition = c(8,8), full.states=TRUE)

#k=c(4,5) is 2 sets of chars, one with 4 states and one with 5 char states
#trait.num=c(8,8) is two sets of 8 chars each
#br.rates = faster than calyx
#full.states=TRUE means all char. states will be included in the simulated trait data

#combine  partitions
combined_crinoid <- combine.morpho(calyx_partition, arms_partition)


#Simulate missing data for each partition; less missing data in calyx partition,
    #more missing data in arms partition
final_crinoid_part <- sim.missing.data(
    data = combined_crinoid, method = "partition",
    seq = "tips", probability = c(0.2, 0.2, 0.5, 0.5)
)
```

### Group 2 : Future model development

Authors: David, April, Mark, Ugne, Davy

-   The tripartite structure of Bayesian time-tree inference could potentially be broken down. Spike clock models introduce a connection between the tree model and the clock model. We could introduce a connection between the tree model and the substitution model as well, e.g. by assuming that the $Q$-matrix changes at speciation events.

-   Discussion of the work April did with Pete Wagner, where the $Q$-matrix itself is time-dependent. The transition probabilities are:

    $$
    P_{i,j} = \frac{1}{k} \left(1 - e^{-k \delta (1 + \lambda t)} \right)
    $$

    $$
    P_{i,i} = \frac{1}{k} \left(1 - e^{-k \delta (1 + \lambda t)} \right) + e^{-k \delta (1 + \lambda t)}
    $$

-   This increases the number of dependencies, but not the number of free parameters! You are already estimating $\lambda$ (speciation rate) as part of the tree model, and $\delta$ (branch rate) as part of the clock model.

-   When this model was first described, it wasn't possible to let the speciation rate vary continuously through time, so the tree was cut up into narrow time slices.

-   Can we pinpoint properties in which character matrices generated under this model differ from those generated under the standard M$k$ model? Maybe use the test statistics from Mulvey et al. (2025)?

-   Can we fit punctuated evolution models to distributions of trees that were inferred under models without punctuation?

-   Discussion of biogeographic dating (FBD + DEC)

-   Discussion of stepping-stone sampling

### Group 3: Impacts of ambiguous characters

Authors: Bethany, Pedro, Melanie, Mariana, Tim, Filippo

Step 1: Biology

We created a simple simulation considering the effect of silicification on brachiopods, where silicification randomly preserves the internal structure of shells. The first partition creates external shell characters, notably whether a taxon has few, moderate, or many costae (ribs). The second partition assigns random internal characters present to reflect whether or not these taxa were silicified.

```{R}
library(TreeSim)
library(MorphSim)
set.seed(12)

tree <- sim.bd.taxa(n = 10, numbsim = 1, lambda = 1, mu = 0.5, complete = TRUE)[[1]]

# Partition 1: external shell, ordered meristic characters (e.g. few / moderate / many ribs)
ord_Q <- matrix(c(-0.5, 0.5, 0,
                  1/3, -2/3, 1/3,
                  0, 0.5, -0.5), nrow = 3, byrow = TRUE)

shell_ext <- sim.morpho(time.tree = tree, k = 3, trait.num = 10, br.rates = 0.1,
                        variable = TRUE, define.Q = ord_Q)

# Partition 2: internal characters, unordered, strongly heterogeneous rates
# (alpha < 1 = some characters nearly frozen, a few fast)
shell_int <- sim.morpho(time.tree = tree, k = 3, trait.num = 10, br.rates = 0.1,
                        variable = TRUE, ACRV = "gamma", alpha.gamma = 0.5, ACRV.ncats = 4)

brach <- combine.morpho(shell_ext, shell_int)
brach$model$Specified


brach_missing <- sim.missing.data(data = brach, method = "partition",
                                  seq = "tips", probability = c(0.1, 0.7))


```

Step 2: Stratigraphy

Silicification is currently thought to be controlled by lithology, such that silicification is more likely to occurr in carbonate rocks with minimal fine-grained siliciclastic content. The amount of terrigenous fine-grained siliciclastic in a carbonate system is controlled by accomodation, whereby any major decrease in accomodation subaerially exposes the platform, decreasing river baselines, and introducing siliciclastic sediment into the basin.

Step 3: Limitations

Many limitations were preliminarily considered. Notably the role of water depth, the intersection of various taphonomic factors, which characters to simulate as potentially losing, and how to include ambigious traits?

### Group 4: Biases in missing data

Authors: Richard, Walker, Madleen, Martin, Will

We are looking at the effect of having biased missing data where you have one set of data that is often found but not phylogenetically uninformative (e.g. teeth) and a second set that has loads of missing data but which is phylogenetically character-rich (e.g. body fossils).

Building the tree

```{r}
set.seed(5)

tree_list <- sim.bd.taxa(
  n        = 30,   # number of extant tips
  numbsim  = 1,    # number of trees to simulate
  lambda   = 1,    # speciation rate
  mu       = 0.8,  # extinction rate
  complete = TRUE # drop extinct/non-sampled lineages; keep only extant tips
)

tree <- tree_list[[1]]

```

Partition 1: Teeth Setting up a tooth partition, fast evolving, few characters

```{r}

morpho_mkv_teeth <- sim.morpho(k = 2, time.tree = tree, br.rates = 0.2,
                          trait.num = 15, variable = TRUE)

```

Partition 2: Braincase Setting up a braincase partition, slow evolving, lots of characters

```{r}

morpho_mkv_brain <- sim.morpho(k = 2, time.tree = tree, br.rates = 0.05,
                          trait.num = 85, variable = TRUE)
```

Combining partitions

```{r}

combined <- combine.morpho(morpho_mkv_brain, morpho_mkv_teeth)

```

Introduce missing data We have made our neurocranial parittion have loads more missing data than the teeth. We would ideally like to be able to do this only introducing missing data for extant tips and don't know how to combine that with doing it by partition. Maybe if partition was a different argument to method then you could combine them.

```{r}
# hard tissue characters (partition 1): 10% missing
# soft tissue characters (partition 2): 70% missing
missing_part <- sim.missing.data(
  data = combined, method = "partition",
  seq = "tips", probability = c(0.95, 0.05)
)
```

Writing files

```{r}
# full tree
#write.morpho(missing_part, file = "Fishtree.tre", type = "tree",
#             reconstructed = FALSE)

# character matrix
#write.morpho(missing_part, file = "Fishmatrix.nex", type = "matrix", reconstructed = FALSE)

# fossil occurrence ages (compatible with RevBayes)
#write.morpho(missing_part, file = "Fishages_rb.tsv", type = "ages")

# ages with associated uncertainty
#write.morpho(missing_part, file = "Fishages.tsv", type = "ages",
#             uncertainty = 2)
```

### Group 5: Testing the Realism of Simulated Morphological Data

Authors: Stephen, Myrthe, Kenneth, Anthony

Adjusted R script of Morphsim to represent a conceptual archosaur phylogeny. An attempt was made to create several partitions with multiple character states and with differences in missing characters (e.g. more in multistate characters).

```{R}
set.seed(12)

tree_list <- sim.bd.taxa(
  n        = 10,   # number of extant tips
  numbsim  = 1,    # number of trees to simulate
  lambda   = 1,    # speciation rate
  mu       = 0.5,  # extinction rate
  complete = TRUE # if set to false drop all extinct lineages and keep only extant tips
)

tree <- tree_list[[1]]
# total tip count, including BOTH extant and extinct lineages
# (this will be > 10, since n only controls extant tips)morpho_mk <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15)


morpho_mk <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15)

morpho_mkv <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1,
                         trait.num = 15, variable = TRUE)
# tolerate autapomorphies
morpho_pars_standard <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15, variable = TRUE, parsimony = "standard")

# every state must occur in 2+ taxa = stricter
morpho_pars_strict <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1,
                                 trait.num = 15, variable = TRUE,
                                 parsimony = "strict")
morpho_acrv <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15, ACRV.ncats = 4)



ord_Q <- matrix(c(-0.5, 0.5, 0.0,
                  0.3333333, -0.6666667, 0.3333333,
                  0.0, 0.5, -0.5),
                nrow = 3, byrow = TRUE)

morpho_ordered <- sim.morpho(k = 3, time.tree = tree, br.rates = 0.1, trait.num = 15, variable = TRUE, ACRV = "gamma", alpha.gamma = 1,
                             ACRV.ncats = 4, define.Q = ord_Q)





# partition 1: unordered binary characters with MkV + gamma ACRV, present/absent, assumed to change less frequently
partition_1 <- sim.morpho(time.tree = tree, k = 2, trait.num = 72,
                          br.rates = 0.1, variable = TRUE,
                          ACRV = "gamma", alpha.gamma = 1, ACRV.ncats =4)

# partition 2: ordered three-state characters with a custom Q matrix
partition_2 <- sim.morpho(time.tree = tree, k = 3, trait.num = 8, variable = TRUE, br.rates = 0.3, define.Q = ord_Q)

# partition 3: unordered three-state characters with MkV + gamma ACRV
partition_3 <- sim.morpho(time.tree = tree, k = 3, trait.num = 13,
                          br.rates = 0.3, variable = TRUE,
                          ACRV = "gamma", alpha.gamma = 1, ACRV.ncats =4)

c1 <- combine.morpho(partition_1, partition_2)
#combined <- combine.morpho(c1, partition_3)


## COMMENTED OUT UNTIL LAURA FIXES BUG ######

# 100 traits total, model info from both partitions preserved
#length(combined$sequences$tips[[1]])
#combined$model$Specified
# partition 1: 30% missing
# partition 2: 70% missing
# partition 3: 70% missing

#missing_part <- sim.missing.data(
#  data = combined, method = "partition",
#  seq = "tips", probability = c(0.3, 0.7, 0.7)
#)

#missing_extinct <- sim.missing.data(
#  data = combined , method = "extinct",
#  seq = "tips", probability = 0.5
#)
```

