---
title: "Time in rocks"
---

## Introduction

This chapter introduces the concept of **age-depth models**, which formalize geochronological knowledge and contain a lot of stratigraphic information, e.g. on sedimentattion rates and hiatus location and duration.

At the end of the chapter you will

* be able to explain what an age-depth model is and what stratigraphic information it contains
* have an intuition on how age-depth models modify the stratigraphic distribution of fossils

These materials are adopted form the workshop "(Modeling Stratigraphic Paleobiology in R)[https://mindthegap-erc.github.io/StratPal_workshop_materials/]".

## Age-depth models

First load the data from your stratigraphic forward model:

```{r}
adm_name = "data/output/columns_adm.csv" # adjust if you have a different output location
data_adm = read.csv(adm_name) # read age-depth model data
names(data_adm) # inspect names

```

Next, extract the time-steps and the associated heights at the examined location. Your results and location heights might differ depending on your simulation setup
```{r}
t =  data_adm[,names(data_adm)[2]] # time step
h_slope = data_adm[,names(data_adm)[3]]
h_margin = data_adm[,names(data_adm)[4]]
h_proximal = data_adm[,names(data_adm)[5]]
h_midplat = data_adm[,names(data_adm)[5]]
```

Now we can start constructing age-depth models:

```{r}
adm_midplat = admtools::tp_to_adm(t = t,
                                  h = h_midplat,
                                  T_unit = "Myr",
                                  L_unit = "m")
```

These age-depth model objects automatically identify hiatuses, and allow convenient extraction of information that is not obvious from the time and height data alone.

Let's start by plotting the age-depth model:

```{r}
plot(adm_midplat,
     lwd_destr = 1, # line width destructive interval (hiatus)
     lwd_acc = 2, # line width accumulative interval
     col_destr = "blue",
     col_acc = "black")
admtools::T_axis_lab() # add time axis label
admtools::L_axis_lab() # add length axis label (stratigraphic domain)
```

### Extracting information from age-depth models

Age-depth models encode a lot of crucial stratigraphic information

```{r}

```


### Points of discussion

1. 
